A geometric framework for capturing high-resolution insights into conceptual knowledge and learning in classroom-like settings

### 2023-01-28 paper/main.tex
first pass of edits through end of intro

The difference between ``understanding'' and ``memorizing,'' as framed by
researchers in education, cognitive psychology, and cognitive
neuroscience~\citep[e.g.,][]{Kato40, Gall00, ScotEtal07, HallGree08, Macl05} has
profound analogs in the fields of natural language processing and natural
language understanding. For example, considering the raw contents of a document
(e.g., its constituent symbols, letters, and words) might provide some
information about what the document is about, just as memorizing a passage
might provide some ability to answer simple questions about it~\citep[e.g., whether
it might contain words related to furniture versus physics;][]{LandDuma97,
BleiEtal03, BleiLaff06}. However, modern natural language processing
models~\citep[e.g.,][]{MikoEtal13a, CerEtal18, BrowEtal20} also attempt to
capture the deeper meaning \textit{underlying} those atomic elements. These
models consider not only the co-occurrences of those elements within and across
documents, but also patterns in how those elements appear across different
scales (e.g., sentences, paragraphs, chapters, etc.), the temporal and
grammatical properties of the elements, and other high-level characteristics of
how they are used~\citep{Mann20, Mann21a}. According to these models, the deep
conceptual meaning of a document may be captured by a feature vector in a
high-dimensional representation space, where nearby vectors reflect
conceptually related documents. A model that succeeds at capturing an analog of
``understanding'' is able to assign nearby feature vectors to two conceptually
related documents, \textit{even when the words contained in those documents
have very little overlap}.

@@

Here we develop a framework for modelling how knowledge is acquired during
learning. The central idea is to use text embedding models to define the
coordinate systems of two maps: (a) a \textit{knowledge map} that describes the
extent to which each concept is currently known and (b) a \textit{learning map}
that describes changes in knowledge over time. Each location on these maps
represents a single concept, and the maps' geometries are defined such that
related concepts are located nearby in space. We use this framework to analyze
and interpret behavioral data collected from an experiment that has
participants watch and answer multiple choice questions about a series of
recorded course lectures.

Our primary research goal is to advance our understanding of what it means to
acquire deep real-world conceptual knowledge. Traditional laboratory approaches
to studying learning and memory (e.g., list learning studies) often draw little
distinction between memorization and understanding. Instead, these studies
typically focus on whether information is effectively encoded or retrieved,
rather than whether the information is \textit{understood}. Approaches to
studying conceptual learning, such as category learning experiments, can start
to investigate the distinction between memorization and understanding, often by
training participants to distinguish arbitrary or random features in otherwise
meaningless categorized stimuli. However the objective of real-world training,
or learning from life experiences more generally, is often to develop new
knowledge that may be applied in \textit{useful} ways in the future. In this
sense, the gap between modern learning theories and modern pedagogical
approaches and classroom learning strategies is enormous: most of our theories
about \textit{how} people learn are inspired by experimental paradigms and
models that have only peripheral relevance to the kinds of learning that
students and teachers actually seek. To help bridge this gap, our study uses
course materials from real online courses to inform, fit, and test models of
real-world conceptual learning. We also provide a ``proof of concept''
demonstration of how our models might be used to construct ``maps'' of what
students know, and how their knowledge changes with training. In addition to
helping to visualize knowledge (and changes in knowledge), we hope that such
maps might lead to real-world tools for improving how we educate.

@@

At its core, our main modeling approach is based around a simple assumption
that we sought to test empirically: all else being equal, knowledge about a
given concept is predictive of knowledge about similar or related concepts.
From a geometric perspective, this assumption implies that knowledge is
fundamentally ``smooth.'' In other words, as one moves through a space
representing someone's knowledge (where similar concepts occupy nearby
coordinates), their ``level of knowledge'' should change relatively gradually
throughout that space. To begin to test this smoothness assumption, we sought
to track our participants' knowledge and how it changed over time in response
to training.

>>>>

The difference between ``understanding'' and ``memorizing,'' as framed by
researchers in education, cognitive psychology, and cognitive
neuroscience~\citep[e.g.,][]{Kato40, Gall00, ScotEtal07, HallGree08, Macl05} has
profound analogs in the fields of natural language processing and natural
language understanding. For example, considering the raw contents of a document
(e.g., its constituent symbols, letters, and words) might provide some
information about what the document is about, just as memorizing a passage
might provide some ability to answer simple questions about it~\citep[e.g., whether
it contains words related to furniture versus physics;][]{LandDuma97,
BleiEtal03, BleiLaff06}. However, modern natural language processing
models~\citep[e.g.,][]{MikoEtal13a, CerEtal18, BrowEtal20} also attempt to
capture the deeper meaning \textit{underlying} those atomic elements. These
models consider not only the co-occurrences of those elements within and across
documents, but also patterns in how those elements appear across different
scales (e.g., sentences, paragraphs, chapters, etc.), the temporal and
grammatical properties of the elements, and other high-level characteristics of
how they are used~\citep{Mann20, Mann21a}. According to these models, the deep
conceptual meaning of a document may be captured by a feature vector in a
high-dimensional representation space, where nearby vectors reflect
conceptually related documents. A model that succeeds at capturing an analog of
``understanding'' is able to assign nearby feature vectors to two conceptually
related documents, \textit{even when the words contained in those documents
have very little overlap}.

@@

Here we develop a framework for modeling how knowledge is acquired during
learning. The central idea is to use text embedding models to define the
coordinate systems of two maps: (a) a \textit{knowledge map} that describes the
extent to which each concept is currently known and (b) a \textit{learning map}
that describes changes in knowledge over time. Each location on these maps
represents a single concept, and the maps' geometries are defined such that
related concepts are located nearby in space. We use this framework to analyze
and interpret behavioral data collected from an experiment that had
participants watch and answer multiple choice questions about a series of
recorded course lectures.

Our primary research goal is to advance our understanding of what it means to
acquire deep real-world conceptual knowledge. Traditional laboratory approaches
to studying learning and memory (e.g., list learning studies) often draw little
distinction between memorization and understanding. Instead, these studies
typically focus on whether information is effectively encoded or retrieved,
rather than whether the information is \textit{understood}. Approaches to
studying conceptual learning, such as category learning experiments, can start
to investigate the distinction between memorization and understanding, often by
training participants to distinguish arbitrary or random features in otherwise
meaningless categorized stimuli. However the objective of real-world training,
or learning from life experiences more generally, is often to develop new
knowledge that may be applied in \textit{useful} ways in the future. In this
sense, the gap between modern learning theories and modern pedagogical
approaches and classroom learning strategies is enormous: most of our theories
about \textit{how} people learn are inspired by experimental paradigms and
models that have only peripheral relevance to the kinds of learning that
students and teachers actually seek \citep{Macl05}. To help bridge this gap, our study uses
course materials from real online courses to inform, fit, and test models of
real-world conceptual learning. We also provide a ``proof of concept''
demonstration of how our models might be used to construct ``maps'' of what
students know, and how their knowledge changes with training. In addition to
helping to visualize knowledge (and changes in knowledge), we hope that such
maps might lead to real-world tools for improving how we educate.

@@

At its core, our main modeling approach is based around a simple assumption
that we sought to test empirically: all else being equal, knowledge about a
given concept is predictive of knowledge about similar or related concepts.
From a geometric perspective, this assumption implies that knowledge is
fundamentally ``smooth.'' In other words, as one moves through a space
representing an individual's knowledge (where similar concepts occupy nearby
coordinates), their ``level of knowledge'' should change relatively gradually
throughout that space. To begin to test this smoothness assumption, we sought
to track our participants' knowledge and how it changed over time in response
to training.
